Snow algae in the alpine and subalpine zones of British Columbia

### 2020-01-07 manuscript/manuscript.Rmd
responded to LQ results comments

We collected 309 snow algae samples from 13 mountains throughout the summer of 2018. We first detected snow algae on May 18, and we continued to sample our low elevation sites at Seymour (S) and Hollyburn (H) until the snow had completely melted out by early July. We did not detect snow algae above treeline until June 20, and snow persisted at higher elevations throughout the entire summer.

We found morphologically distinct snow algae communities in different habitats. In shaded, forested sites we often found green or orange patches of snow, containing green or orange ovate cells with short spines resembling *Chloromonas cf. brevispina* (Fig. 1c). In forest clearings we observed low-angle linear drainage channels or runnels containing high concentrations of snow algae. In May and early June these runnels contained green snow hidden 2 to 5 cm below the white snow surface, and on subsequent visits these same sites contained orange snow at the surface, with morphologies resembling *Chloromonas krienitzii* (Fig 1b). Red snow was prevalent above treeline, dominated by red cells similar in appearance to *Sanguina nivaloides*, although we also occasionally observed green snow patches above treeline. We found red snow dominated by *Chlainomonas rubra* morphologies in unshaded sites at all elevations.

We chose 33 snow algae samples for 18S and *rbcL* amplicon Illumina sequencing. Both libraries were dominated by amplicon sequence variants (ASVs) that were assigned to Chlorophyta. Our 18S library contained 67 ASVs that were assigned to Chlorophyta and 7 ASVs to Ochrophyta, while our *rbcL* library contained 642 ASVs, of which 603 were assigned to Chlorophyta. The top genera were *Chloromonas*, *Chlainomonas*, and *Sanguina*, but most ASVs were not assigned to genus level (Fig. 2). We used t-SNE to generate operational taxonomic units (OTUs), and found many OTUs that did not correspond to any known taxa on GenBank: several OTUs related to *Chloromonas*, and one Chlamydomonadaceae OTU that was only assigned to the species level.

Above and below treeline communities contained some unique taxa, but most taxa did not exhibit a clear elevational trend (Fig. 3). *Sanguina* was the dominant taxa in most samples taken from above treeline. Additionally, high elevation sites contained several taxa not found at low elevations: an unannotated OTU of *Chloromonas*, and ASVs with a best BLAST match to *Raphidonema longiseta* (KM462868.1) were found at the three highest elevation sites. Green snow from alpine sites contained *Chloromonas* OTUs that were also found in sites below treeline. Low elevation sites contained *Chloromonas krienitzii*, but many sites were dominated by taxa that were found at all elevations, such as *Chlainomonas*.

Cell counts, 18S, and *rbcL* datasets were generally consistent, but we did notice some discrepancies. *Chlainomonas* was more prevalent in our *rbcL* library than in cell counts, while *Sanguina* was underrepresented in *rbcL* compared to 18S and cell counts. We did not detect *Chlainomonas* in 18S.

At runnel sites on Hollyburn and Seymour we observed snow algae blooms that changed from green to orange over the course of the summer (Fig. 1). We sequenced both green subsurface and orange surface snow at three different sites, and found both were dominated by *Chloromonas krienitzii*, although surface samples also contained some *Chlainomonas* (Supplementary _).

>>>>

We collected 309 snow algae samples from 13 mountains throughout the summer of 2018. At our low elevation sites (850 - 1500 m) we detected snow algae as early as May 18, but these sites were also the earliest to melt out. We did not detect snow algae at high elevation sites (1500 - 2200 m) until June 20, where both the snow and snow algae blooms persisted throughout the entire summer.

Using light microscopy we identified morphologically distinct snow algae in different habitats (Fig. 1, S4). Red snow was prevalent in areas of high solar exposure; above treeline sites were often dominated by red cells similar in appearance to *Sanguina nivaloides*[@prochazkova_sanguina_2019] (Fig. 1a,b). We found cells resembling *Chlainomonas rubra*[@novis_two_2008] at all elevations, in highest relative abundance in areas recieving direct sunlight. We occasionally observed green snow patches in alpine meadows recieving full sun, but more frequently observed green snow in well-shaded, forested sites. At four low-angle sites in clearings near treeline we observed snow melt channels or runnels containing high concentrations of snow algae (Fig. 1c). In May and early June these sites contained green snow hidden 2 to 5 cm below the white snow surface, but upon subsequent visits the same location contained orange snow at the surface, whose morphologies resembled published images of *Chloromonas krienitzii*[@matsuzaki_revisiting_2015] (Fig. 1d). Forested sites at low elevations often contained green or orange patches of snow, containing green or orange ovate cells with short spines resembling *Chloromonas cf. brevispina*[@matsuzaki_revisiting_2015] (Fig. 1e,f), although we also occasionally observed red snow at forested sites containing morphologies similar to *Sanguina nivaloides*.

Of the 309 samples studied by light microscopy, we selected 33 snow algae samples for 18S and *rbcL* amplicon Illumina sequencing. Both libraries were dominated by amplicon sequence variants (ASVs) that were assigned to Chlorophyta. Our 18S library contained 67 ASVs that were assigned to Chlorophyta and 7 ASVs to Ochrophyta, while our *rbcL* library contained 642 ASVs, of which 603 were assigned to Chlorophyta, the remaining 41 were Trebouxiophyceae. The top genera assigned by the taxonomy assignment algorithm IDTaxa[@murali_idtaxa_2018] were *Chloromonas*, *Chlainomonas*, and *Sanguina*; although most ASVs were not assigned to genus level by this algorithm, the majority of ASVs clustered near one of these genera in NMDS ordination (Fig. 2a). We used t-SNE to generate operational taxonomic units (OTUs), and found that many OTUs that did not correspond to any known taxa on GenBank (Fig. 2b). These included three OTUs that were closely related to *Chloromonas*, and one Chlamydomonadaceae OTU that was only assigned to the family level, meaning that it did not match any known algae genus on GenBank.

Above and below treeline samples contained some unique taxa, but most taxa did not exhibit a clear elevational trend (Fig. 3). *Sanguina* was the dominant taxa in most samples from above treeline. Additionally, high elevation sites contained several taxa not found at low elevations: an unannotated OTU of *Chloromonas*, and ASVs with a best BLAST match to *Raphidonema longiseta* (KM462868.1) were found at the three highest elevation sites. Green snow from high alpine sites contained *Chloromonas* OTUs that were also found in sites below treeline. Low elevation sites contained *Chloromonas krienitzii*, but many sites were dominated by taxa that were found at all elevations, such as *Chlainomonas*.

Cell counts, 18S, and *rbcL* datasets were generally consistent, but there were some interesting discrepancies. *Chlainomonas* was more prevalent in our *rbcL* library than in cell counts, while *Sanguina* was underrepresented in *rbcL* compared to 18S and cell counts. We did not detect *Chlainomonas* with 18S sequencing.

At runnel sites on Hollyburn and Seymour we observed snow algae blooms that changed from green to orange over the course of the summer (Fig. 1). We sequenced both green subsurface and orange surface snow at three different sites, and found both were dominated by *Chloromonas krienitzii*, although surface samples also contained *Chlainomonas* (Supplementary __).
